### 2016-07-31 doc/publications/plos16/plos16.tex
edits to CursedForest.bib plos16.tex

\section*{Notes}
\begin{itemize}
\item what are you guys editing this with?  what about some carriage
  returns or line feeds or something!
\item Are we going to use the TCGA data in this paper?
\item are we going to run sparkML or Cursed forest on the extended
  simulation data \cite[]{Genuer.et.al.2010}? Has Aidan done this already?
\item I dont think I can get R to handle chromosme 1 etc. May have to run CF on chromosme 22, see table
  \ref{table:smaller_data} -- On it, Aidan
\item check Piotr's simulation
\item run an argument as follows:
  \begin{itemize}
  \item variable  selection may succeed where accurate prediction does not
  \item consider  \cite{Segal.2004} examples
  \item  perhaps we argue that in the case where the relationship of $X$ and $y$ takes a particular form we can recover
    the reltionship for very large $p$p
  \end{itemize}
\item \cite{Tuv.et.al.2009} feature selection via adding permuted variables. Do we want to consider this?
\item chek out varSelRF
\end{itemize}

@@

The digital revolution or `datafication' sees a dramatic increase in data collected about every aspect of   \marginpar{is seeing, will see}
life~\cite{Loebbecke2015}.  These dataset not only grow vertically by capturing more and more events but also
horizontally by capturing more information about these events.  The challenge of big and `wide' data is especially
pronounced in the health space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3
billion basepairs of the human genome.  While a large fraction of this data is identical between individuals, \marginpar{base-pairs}
identifying the relevant and disease specific differences is the focus of complex trait analysis [citation].

@@

% Developed b Apache, Hadoop is one platform for dealing with large data sets is Platforms Fortunately computational
% tools such as Apache Spark are available as a framework for iteratively dealing with big data.
Distributing the compute tasks is one strategy to overcome this challenge. Apache Spark, based on Scala, allows for  
\marginpar{is it based on scale or does it have a scale API?}
distributed applications to scale to thousands of nodes.  A Spark application runs on a `driver' node, and
this node divides and distributes work out to the many `worker' nodes.  Spark also includes the machine learning (ML)
library, Spark ML, providing popular machine learning libraries.  Spark ML includes supervised approaches such as
decision trees and the ensemble method, random forests, as well as unsupervised approaches such as k-means clustering.

@@

While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is
supervised learning, i.e. given a phenotype of clinical label determine the most associated features.  One of the
simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a
linear combination of the feature variables.  Processing the `wide' genomic data poses difficulties for machine learning
applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
This is known as the `curse of dimensionality'~\cite{Bauer2014}.  
\marginpar{is this the reference we want for the `curse of dimensionality'}

Decision trees have a number of desirable features, they:
\begin{itemize}
  \item are scale invariant;
  \item can handle categorical and real valued inputs;
  \item can handle missing values;
  \item are insensitive to useless predictors;
  \item are able to capture interactions between features (which is of importance for modelling complex polygenic diseases).
  \end{itemize}
However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
lead to a very different model. 
\marginpar{a potential problem  with tree models is the tendencey to split on categorical variables with a high number of
  levels, even when better splits are available}
\subsection*{RF}

@@

The results are combined in an appropriate way, i.e. for a regression problem,
\begin{equation*}
{\bar {h}}={\frac {1}{B}}\sum _{b=1}^{B} h(x;\theta_b).
\end{equation*}
\marginpar{make it a classification example}
%where $x$ is the covariate vector of length $p$
%with associated random vector $X$ and the $\theta_b$ are i.i.d. random vectors.

@@

\subsection*{benefits of RF}

\marginpar{weak learners only in a misuse of the term. ``weak learners'' was applied to decision stumps}
%http://stats.stackexchange.com/questions/77018/random-forest-is-it-a-boosting-algorithm
%http://stats.stackexchange.com/questions/173390/gradient-boosting-tree-vs-random-forest
%overcomes this issue and is hence ideally suited to this application.  Furthermore, RF are able to capture interactions
%between features, which is of importance for modelling complex polygenic diseases.

@@

Random forests are a variance reduction technique. They take a tree models which have a low bias but a high variance and
by combining them reduce the variance. The prediction error is the sum of the variance and the bias squared.

The benefits of random forests models are
  \begin{itemize}
  \item they give an out-of-bag estimate of model accurace;
  \item they provide a measure of variable importance;
  \item they inherit many of the qualities of tree models.
\marginpar{but not the handling of missing velues via surrogate variables. This would be difficult as the trees as split
  on random subsets of variables}
  \end{itemize}

@@

\subsection*{Random Forests converges on wide data}

As discussed in the Introduction,
\marginpar{I think this has been removed from the introduction}
random forests are resistent against overfitting.
However it is not true that they will not overfit; see \cite{Segal.2004} where examples of RF overfitting are given 
\marginpar{discuss these examples}

\cite{Strobl.et.al.2007} demosnstrate that a RF algorithm is subject to a bias in variable selection via two mechanisms:
\begin{itemize}
\item variable scale (for $X_I$ continuous) and number of levels (for $X_i$ discrete) inroduces a bias
\item boosting introduces a bias.
\end{itemize}
These effects will be apparent in the CF algorithm as well. \marginpar{we can fix the boosting effect}
However, there are many examples, (particularly in genomics) whre we have data sets that are
\begin{itemize}
\item very wide
\item all variables have the same number of levels (measures at different points along the genome)
\item the hypothesied functional relationship is of a form amenable to a DT.
\end{itemize}

@@

\subsection*{Parameter settings}
\marginpar{ was''Random forest is sensitive to tree depth''}
In this setting, variable selection in very high dimensions, we nned to consider the parameter setting for the RF
algorithm.

I will use the R terms from the package RandomForests whcih incoporates the original Fortran code by Brieman and
Cutler. We consider the parameters:
\begin{itemize}
\item the size and complexity of the individual trees is controlled in RandomForests by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark.ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest. The SparkML documentation somewhat unhelpfully sets \texttt{maxDepth=4} in their classification
  example\footnote{the documentation is available at \url{http://spark.apache.org/docs/latest/mllib-ensembles.html} }.
  In the ATCG example the RF code produced trees with depths of 30+. We note that Random forests are a way of averaging
  multiple deep decision trees, trained on different parts of the same training set, with the goal of reducing the
  variance. \url{https://en.wikipedia.org/wiki/Random_forest} and refs therein.  
\marginpar{are we including this  example?  Supplementary info?}
\item \texttt{ntree} the number of trees 
\item \texttt{mtry} --  the number of variables considered at each split (if \texttt{mtry}=$p$, we have a boosted decison
  tree model \marginpar{is this right?}
\end{itemize}

@@

>>>>

\section*{Notes}
\begin{itemize}
\item what are you guys editing this with?  what about some carriage
  returns or line feeds or something!
\item Are we going to use the TCGA data in this paper?
\item are we going to run sparkML or Cursed forest on the extended
  simulation data \cite[]{Genuer.et.al.2010}? Has Aidan done this already?
\item I dont think I can get R to handle chromosme 1 etc. May have to run CF on chromosme 22, see table
  \ref{table:smaller_data} -- On it, Aidan
\item check Piotr's simulation
\item run an argument as follows:
  \begin{itemize}
  \item variable  selection may succeed where accurate prediction does not
  \item consider  \cite{Segal.2004} examples
  \item  perhaps we argue that in the case where the relationship of $X$ and $y$ takes a particular form we can recover
    the reltionship for very large $p$p
  \end{itemize}
\item \cite{Tuv.et.al.2009} feature selection via adding permuted variables. Do we want to consider this?
\item check out varSelRF
I think the structure of the first section should be
\begin{itemize}
\item  convergence results from Biau 2012
\item simulations reproduced from Genuer et al 2010
\item we extend the simulations to many more variables
\item discussion of the effect of the ntree and mtry parameters for variable selection with many noise variables
\item  in the case of some functions (step, ramp) we can recover the significant variables in the presence of a very large number of noise variables
\item in the case of genomic data, our functions of interest are more likely to be stop or ramp functions rather than harmonic functions. So we have some hope.
\end{itemize}

\item check out 
  \begin{itemize}
  \item D\'iaz-Uriarte, R., \& Alvarez de Andr\'es, S. (2006). Gene selection and classification of microarray data using random
    forest. BMC Bioinformatics, 7, 3. http://doi.org/10.1186/1471-2105-7-3
  \item Goldstein et al. (2011)
  \item Chen (2012)
  \item VSURF and  varSelRF
  \end{itemize}
\end{itemize}

@@

The digital revolution is seeing a dramatic increase in data collected about every aspect of 
life~\cite{Loebbecke2015}.  These dataset not only grow vertically by capturing more and more events but also
horizontally by capturing more information about these events.  The challenge of big and `wide' data is especially
pronounced in the health space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3
billion basepairs of the human genome.  While a large fraction of this data is identical between individuals, \marginpar{base-pairs}
identifying the relevant and disease specific differences is the focus of complex trait analysis [citation].

@@

% Developed b Apache, Hadoop is one platform for dealing with large data sets is Platforms Fortunately computational
% tools such as Apache Spark are available as a framework for iteratively dealing with big data.
Distributing the compute tasks is one strategy to overcome this challenge. Apache Spark, based on Scala, allows for  
\marginpar{is it based on Scala or does it have a Scala API?}
distributed applications to scale to thousands of nodes.  A Spark application runs on a `driver' node, and
this node divides and distributes work out to the many `worker' nodes.  Spark also includes the machine learning (ML)
library, Spark ML, providing popular machine learning libraries.  Spark ML includes supervised approaches such as
decision trees and the ensemble method, random forests, as well as unsupervised approaches such as k-means clustering.

@@

While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is
supervised learning, i.e. given a phenotype of clinical label determine the most associated features.  One of the
simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a
linear combination of the feature variables.  Processing the `wide' genomic data poses difficulties for machine learning
applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
This is known as the `curse of dimensionality'~\cite{Bauer2014}.  
\marginpar{is this the reference we want for the ``curse of dimensionality''  Bellman (1961) is a more ususl reference}

Decision trees have a number of desirable features, they:
\begin{itemize}
  \item are scale invariant;
  \item can handle categorical and real valued inputs;
  \item can handle missing values;
  \item are insensitive to useless predictors;
  \item are able to capture interactions between features (which is of importance for modelling complex polygenic diseases).
  \end{itemize}
However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
lead to a very different model.

\subsection*{RF}

@@

The results are combined in an appropriate way. In a classification problem we take the majority vote for the trees over
the $i=1,..,Q$ classes
\begin{equation*}
{\bar {h}}=  \arg \max_i \left(\sum_b I(h(x;\theta_b)=i)\right).
\end{equation*}

Random forests are a variance reduction technique. They take a tree models, which have a low bias but a high variance, and
by combining them reduce the variance. The prediction error is the sum of the variance and the bias squared.

@@

%http://stats.stackexchange.com/questions/77018/random-forest-is-it-a-boosting-algorithm
 %http://stats.stackexchange.com/questions/173390/gradient-boosting-tree-vs-random-forest
%overcomes this issue and is hence ideally suited to this application.  Furthermore, RF are able to capture interactions
%between features, which is of importance for modelling complex polygenic diseases.

@@

The benefits of random forests models are
  \begin{itemize}
  \item they give an out-of-bag estimate of model accurace;
  \item they provide a measure of variable importance;
  \item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
    variables. This would be difficult as the trees as split on random subsets of variables).
  \end{itemize}

Since the original implmentation of random forests there have been a number of developments leading to a better understanding of
the algorithm. 
\begin{itemize}
\item random forests are resistent against overfitting. However it is not true that they will not overfit; see \cite{Segal.2004}
  where an example of RF overfitting is  given, caused by the fact that successive trees were correlated;
\item \cite{Strobl.et.al.2007} demonstrates that the RF algorithm is subject to a bias in variable selection via two mechanisms:
  \begin{enumerate}
  \item variable scale (for $X_I$ continuous) and number of levels (for $X_i$ discrete) inroduces a bias. A uninformative  variable with a large
    number of levels may be selected over a more informative  variable with fewer levels;
  \item bagging introduces a bias, but sampleing about $(1- 1/e) \approx 0.632$  of the data without 
  \end{enumerate}
\end{itemize}

\cite{Strobl.et.al.2007} introoduce changes to the RF algorithm to fix 1) and 2). 
However the level effect will be apparent in the CF algorithm as well. \marginpar{we can fix the boosting effect}
However, there are many examples, (particularly in genomics) whre we have data sets that are
\begin{itemize}
\item very wide
\item all variables have the same number of levels (measures at different points along the genome)
\item the hypothesied functional relationship is of a form amenable to a DT.
\end{itemize}

@@

@@

\subsection*{Parameter settings}
We consider the parameter setting for the RF algorithm. We use the R terms from the package RandomForests
\cite{Liaw.and.Weiner.2002} which incoporates the original Fortran code by Brieman and Cutler. We incprporate the 
advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience.

\begin{itemize}
\item \texttt{ntree} the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
  \marginpar{can we do proxmiity?}  \cite{Liaw.and.Weiner.2002} suggest that a high \texttt{ntree} is necessary to get stable
  estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
  the ranking of the importances is quite stable.  In addition, we note that it is possible for a RF model to have a very poor
  fit, and be useless for prediction, and still have a useful ranking of variable importance.
\item \texttt{mtry} --  the number of variables considered at each split (if \texttt{mtry}=$p$, we have a boosted decison
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \texttt{mtry} may give
  better performance.
\item  For classification problems where the class frequencies are extremely unbalanced it may be necessary to change the prediction
  rule \marginpar{can we do this}
\item the size and complexity of the individual trees is controlled in RandomForests by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark.ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest. The SparkML documentation somewhat unhelpfully sets \texttt{maxDepth=4} in their classification
  example\footnote{The documentation is available at \url{http://spark.apache.org/docs/latest/mllib-ensembles.html}.}.

  In the ATCG example the RandomForests code produced trees with depths of 30+. We note that Random forests are a way of averaging
  multiple deep decision trees, trained on different parts of the same training set, with the goal of reducing the
  variance. \url{https://en.wikipedia.org/wiki/Random_forest} and refs therein.  
\marginpar{are we including this  example?  Supplementary info?}
\end{itemize}

@@

  \item D\iaz-Uriarte, R., \& Alvarez de Andr\es, S. (2006). Gene selection and classification of microarray data using random
    forest. BMC Bioinformatics, 7, 3. http://doi.org/10.1186/1471-2105-7-3
  \item Goldstein, B. A., Polley, E. C., & Briggs, F. B. S. (2011). Random Forests for Genetic Association Studies. Statistical
    Applications in Genetics and Molecular Biology, 10(1), 32. http://doi.org/10.2202/1544-6115.1691

@article{Chen2012323,
title = "Random forests for genomic data analysis ",
journal = "Genomics ",
volume = "99",
number = "6",
pages = "323 - 329",
year = "2012",
note = "",
issn = "0888-7543",
doi = "http://dx.doi.org/10.1016/j.ygeno.2012.04.003",
url = "http://www.sciencedirect.com/science/article/pii/S0888754312000626",
author = "Xi Chen and Hemant Ishwaran",
keywords = "Random forests",
keywords = "Random survival forests",
keywords = "Classification",
keywords = "Prediction",
keywords = "Variable selection",
keywords = "Genomic data analysis ",
abstract = "Random forests (RF) is a popular tree-based ensemble machine learning tool that is highly data adaptive, applies to “large p, small n” problems, and is able to account for correlation as well as interactions among features. This makes \{RF\} particularly appealing for high-dimensional genomic data analysis. In this article, we systematically review the applications and recent progresses of \{RF\} for genomic data, including prediction and classification, variable selection, pathway analysis, genetic association and epistasis detection, and unsupervised learning. "
}
